---
title       : Illustrating the extent of Arctic sea-ice
subtitle    : data from late 1978 to early 2013
author      : Developing Data Products
job         : Coursera Data Science specialization
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow      # 
widgets     : []            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
knit        : slidify::knit2slides
---

<!-- change the greenish title slide to blueish (for ice) -->
<style>
.title-slide {
  background-color: #CCE5FF;
}
.title-slide hgroup > h1, 
.title-slide hgroup > h2 {
  color: #004C99;
}
</style>

## Background

US [National Snow and Ice Data Center](http://nsidc.org/) measures the extent of Arctic sea-ice. [The project app](https://johanna.shinyapps.io/Project) aims to illustrate the diminishing trend in Arctic sea-ice by letting the user explore the measurements available at UsingR package in a dashboard-like fashion.

The application supports three types of graphics that are all demonstrated on the following slides.

1. an annual timeseries as boxplot 
2. seasonal variation as monthly boxplot
3. seasonal variations as lines for each year with a selected year highlighted in red

The user can choose the range of years pictured.

[R-bloggers](http://www.r-bloggers.com/arctic-sea-ice-at-lowest-levels-since-observations-began/) inspired the line chart with a highlighted year.

--- 

## Annual time series shows a diminishing trend

```{r echo = FALSE, fig.width = 14}
library(UsingR)
library(ggplot2)
library(scales)
# read data and format dates
data(nisdc)
nisdc$year <- factor(nisdc$Year)
# use fixed y limits on all graphics to accentuate changes
limits <- c(floor(10* min(nisdc$Extent))/10 - 0.1,
            ceil(10*max(nisdc$Extent))/10 + 0.1)
p <- ggplot(nisdc, aes(year, Extent)) + geom_boxplot() +
    xlab("Year") + ggtitle("Arctic sea-ice by year") +
    ylab("Extent in millions of square kilometers") + ylim(limits) +
    theme(axis.text.x = element_text(angle = 90))
print(p)
```

---- 

## Monthly boxplot illustrates the seasonal variation

```{r echo = FALSE, fig.width = 14}
# months as Jan ... Dec
nisdc$mon <- factor(format(strptime(paste(nisdc$Day, nisdc$Month,
                                          nisdc$Year), "%d %m %Y"),
                           "%b"), levels = month.abb)
start <- 1980
end <- 1999
dataset <- nisdc[nisdc$Year >= start & nisdc$Year <= end, ]
p <- ggplot(dataset, aes(mon, Extent)) + geom_boxplot() + xlab("Month") +
    ggtitle(paste0("Seasonal variation in ", start, "-", end)) +
    ylab("Extent in millions of square kilometers") + ylim(limits)
print(p)
```

--- &twocol

## Line chart with a highlighted year combines the two

*** =left
```{r echo = FALSE, warning = FALSE}
nisdc$year <- factor(nisdc$Year)
# dates as 01-Jan ... 31-Dec
nisdc$date <- as.Date(paste(nisdc$Day, nisdc$Month), "%d %m")
start <- 1979
end <- 1988
dataset <- nisdc[nisdc$Year >= start & nisdc$Year <= end, ]
highlight <- 2012
selected <- nisdc[nisdc$Year == highlight, ]
p <- ggplot(dataset, aes(date, Extent, group = year)) + geom_line() +
    geom_line(data = selected, colour = "red", size = 2) +
    scale_x_date(labels = date_format("%d-%b")) + xlab("") + 
    ggtitle(paste0("Seasonal variation in ", start, "-", end, " with ", 
                   highlight, " highlighted in red")) +
    ylab("Extent in millions of square kilometers") + ylim(limits)
print(p)
```

*** =right

```{r echo = FALSE, warning = FALSE}
nisdc$year <- factor(nisdc$Year)
# dates as 01-Jan ... 31-Dec
nisdc$date <- as.Date(paste(nisdc$Day, nisdc$Month), "%d %m")
start <- 2003
end <- 2012
dataset <- nisdc[nisdc$Year >= start & nisdc$Year <= end, ]
highlight <- 1979
selected <- nisdc[nisdc$Year == highlight, ]
p <- ggplot(dataset, aes(date, Extent, group = year)) + geom_line() +
    geom_line(data = selected, colour = "red", size = 2) +
    scale_x_date(labels = date_format("%d-%b")) + xlab("") + 
    ggtitle(paste0("Seasonal variation in ", start, "-", end, " with ", 
                   highlight, " highlighted in red")) +
    ylab("Extent in millions of square kilometers") + ylim(limits)
print(p)
```

*** =fullwidth

Credits for the two-column layout go to [Zev Ross](http://zevross.com/blog/2014/11/19/creating-elegant-html-presentations-that-feature-r-code/).
